A sheaf--function dictionary for algebraic tori over local fields

### 2013-09-24 shanghai.tex
Added SGA5l changes to the first section.

%\input{introduction}
%In his celebrated thesis, John Tate writes 
%\begin{quotation}
%    {\it Concerning the characters of $k^*$ [the
%    multiplicative group of a $\mathfrak{p}$-adic field], the
%    situation is different from that of $k^+$.  First of all, we are
%    interested in continuous multiplicative maps $\alpha \mapsto
%    c(\alpha)$ of $k^*$ into the complex numbers, not only in the
%    bounded ones, and shall call such a map a quasi-character,
%    reserving the word ``character'' for the conventional character of
%    absolute value $1$. Secondly, we shall find no model for the group
%    of quasi-characters, or even for the group of characters, though
%    such a model would be of the utmost importance.}
%  \cite{tate:67a}*{\S 2.3}
%\end{quotation}
%\noindent
In this paper we find a dictionary between quasicharacters of algebraic tori over local fields and quasicharacter sheaves. 
%
What is a quasicharacter sheaf?
Let $G$ be a smooth, commutative group scheme locally of finite
type over a finite field $\Fq$; $G$ need not be reductive, connected, or of
finite type over $\Fq$. Fix a prime $\ell$ invertible in $\Fq$
and write $m : \GxG \to G$ for the multiplication morphism,
$\bG$ for the base change of $G$ to $\bFq$ and
$\Frob{G} : \bG \to \bG$ for the Frobenius automorphism.
A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank-1 and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms;
$\mu$ and $\phi$ are required to satisfy natural conditions explained below.

@@

We fix a prime $\ell$, invertible in $\Fq$.
For any group $A$, we will denote by $A^*$ the dual group $\Hom_\text{grp}(A, \EEx)$.

@@

\begin{definition}\label{def:QC}
A \emph{quasicharacter sheaf on $G$} is a triple
$\qcs{L}\ceq (\gqcs{L},\mu,\phi)$ where:
\begin{enumerate}
\labitem{(CS.0)}{CS.0} $\gqcs{L}$ is a rank one $\ell$-adic local system on $\bG$
-- a constructible $\ell$-adic sheaf on $\bG$, locally constant on each connected component; 
\labitem{(CS.1)}{CS.1} $\mu: \bm^* \gqcs{L} \to \LxL$ \David{We should explain the $\boxtimes$ notation}
  is an isomorphism of sheaves on $\bGxG$ such that the following diagram commutes, 
  where $m_3 \ceq m\circ (m\tight{1}{\times}{2}\id) = m\circ (\id\tight{2}{\times}{1} m)$;
%
  \[
  \begin{tikzcd}[row sep=30]
  \bm_3^*\gqcs{L} \arrow{rr}{(\bm \tight{1}{\times}{2} \id)^*\mu} \arrow[swap]{d}{(\id \tight{2}{\times}{1} \bm)^*\mu}
    &&  \bm^*\gqcs{L} \boxtimes \gqcs{L} \dar{\mu \tight{0}{\boxtimes}{1} \id} \\
    \gqcs{L} \boxtimes \bm^* \gqcs{L} \arrow{rr}{\id \boxtimes \mu}
    &&  \gqcs{L} \tight{0}{\boxtimes}{0} \LxL
  \end{tikzcd}
  \]
%
\labitem{(CS.2)}{CS.2} $\phi : \Frob{G}^* \gqcs{L} \to \gqcs{L}$ is an
  isomorphism of constructible $\ell$-adic sheaves on $\bG$ compatible with $\mu$ in the sense that the following diagram commutes.
%
  \[
  \begin{tikzcd}[row sep=20]
  \Frob{\GxxG}^* \bm^* \gqcs{L} \arrow{rr}{\Frob{\GxxG}^*\mu}
    && \Frob{\GxxG}^*(\LxL)\\
    \arrow[equal]{u} \bm^*  \Frob{G}^* \gqcs{L} \arrow[swap]{d}{\bm^* \phi}
    && \Frob{G}^*\gqcs{L}\boxtimes \Frob{G}^*\gqcs{L} \dar{\phi\boxtimes \phi} \arrow[equal]{u} \\
    \bm^*\gqcs{L} \arrow{rr}{\mu}
    && \LxL
  \end{tikzcd}
  \]
\end{enumerate}
\end{definition}

@@

\begin{proof}
  Let $\qcs{L}$ be a quasicharacter sheaf on $G$. We start by showing that
  $\bf^*\gqcs{L}$ is a local system of rank~$1$. Let $c_j :\bH^j \hookrightarrow \bH$ be any
  connected component and let $d_i : \bG^i \hookrightarrow \bG$ be the image of $j$ under the group
  homomorphism $\pi_0(\bf) : \pi_0(\bH) \to \pi_0(\bG)$.  Write $\bf^j : \bH^j \to \bG^i$
  for the restriction of $f$ to $\bH^j$.  Since $\gqcs{L}$ is locally constant on the connected
  components of $\bG$, there is a finite \'etale covering
  $a_i : X_i \to \bG^i$ such that $a_i^* (\gqcs{L}\vert_{\bG^i})$
  is constant.  Define $Y_j$ in the following diagram through pullback:
  \[
  \begin{tikzcd}[row sep=20, column sep=30]
   Y_j \dar[dashed]{g_j} \rar[dashed]{b_j} & \bH^j \dar{\bf^j} \rar[hook]{c_j} & \bH \dar{\bf} \\
   X_i \rar{a_i} & \bG^i \rar[hook]{d_i} & \bG
  \end{tikzcd}
  \]
  Then $b_j$ is a finite \'etale covering
  of $\bH^j$ and
  \[
  b_j^* \left( (\bf^* \gqcs{L})\vert_{\bH^j}\right)
  = (\bf\circ c_j \circ b_j)^*\gqcs{L}
  = (d_i\circ a_i\circ g_j)^*\gqcs{L} = g_j^* a_i^* (\gqcs{L}\vert_{\bG^i}).
  \]

  Since $a_i^* (\gqcs{L}\vert_{\bG^i})$ is a constant sheaf, its pullback
  $b_j^* \left( (\bf^*\gqcs{L})\vert_{\bH^j}\right)$ along $g_j$ is also constant.
  Thus $\bf^*\gqcs{L}$ is locally constant on
  $\bH^j$. Since $(\bf^*\gqcs{L})_{\bg} = \gqcs{L}_{f(\bg)}$ for every
  geometric point ${\bg}$ on $\bG$, and since $\gqcs{L}$ has
  rank $1$ on connected components of $\bG$, it follows that
  $f^*\gqcs{L}$ also has rank $1$ on connected components of $\bG$.
  \David{can't we just say rank 1?  Why ``on connected components of $G$?"}
  This completes the proof that $\bf^*\gqcs{L}$ satisfies condition \ref{CS.0}.

@@

\begin{definition}
Let $\QCb(G)$ be the category of pairs $(\qcs{L}_0,\mu_0)$ 
where $\qcs{L}_0$ an $\ell$-adic local system on $G$ of rank one, 
equipped with an isomorphism $\mu_0 : m^* \qcs{L}_0 \to \qcs{L}_0 \boxtimes \qcs{L}_0$ 
satisfying the analogue of Condition~\ref{CS.1} on $G$; 
morphisms in $\QCb(G)$ are defined as in (the second part of) Condition~\ref{CS.2}. 
This is the category of \emph{bounded quasicharacter sheaves} on $G$. 
\end{definition}

@@

With Theorem~\ref{thm:snake} in hand, 
we can see why neither bounded quasicharacter sheaves 
nor finite quasicharacter sheaves are up to the task of 
geometrizing characters of $G(\Fq)$ in the generality we need.  
%
Consider the case when $G$ is the discrete \'etale group scheme $\ZZ$. 
Let $\chi : G(\Fq) \to \EEx$ be the character determined by $\chi(1) = \ell$ 
and let $\qcs{L}$ be a quasicharacter sheaf in the isomorphism class 
corresponding to the character $\chi : G(\Fq) \to \EEx$ under Theorem~\ref{thm:snake}. 
Then $\trFrob{\qcs{L}}(1) = \ell$. Suppose $\qcs{L}$ was a bounded quasicharacter sheaf. 
Then the restriction of $\qcs{L}$ to the connected component $1\in G$ 
would be a $\ell$-adic local system on $1$ of rank-1, 
thus an $\ell$-adic character $\rho_1 : \pi_1(1,{\bar 1}) \to \EEx$ with
$\rho_1(\Frob{\Fq}) = \ell$. But $\pi_1(1,{\bar 1}) = \Gal(\bFq/\Fq) \iso {\hat \ZZ}$ is compact, 
while $\rho_1(\pi_1(1,{\bar 1}))$ is not compact in $\EEx$ (it contains $\ell^\ZZ$). 
So the Weil sheaf $\gqcs{L}$ does not descend to a local system on $G$.
It follows that the quasicharacter sheaf $\qcs{L}$ is not a bounded quasicharacter sheaf.
%
We can play the same game with the character $\chi : G(\Fq) \to \EEx$ 
defined by $\chi(1) = 1+\ell$. In this case, the corresponding quasicharacter sheaf $\qcs{L}$ is bounded,
defined at $1$ by a pro-\'etale cover of $1$, but not finite, since $1+\ell$ is not a root of unity. 
%
These simple examples illustrate why the categories $\QCb(G)$ and $\QCf(G)$ are inadequate to geometrize characters of $G(\Fq)$.

>>>>

%\input{introduction}
%In his celebrated thesis, John Tate writes 
%\begin{quotation}
%    {\it Concerning the characters of $k^*$ [the
%    multiplicative group of a $\mathfrak{p}$-adic field], the
%    situation is different from that of $k^+$.  First of all, we are
%    interested in continuous multiplicative maps $\alpha \mapsto
%    c(\alpha)$ of $k^*$ into the complex numbers, not only in the
%    bounded ones, and shall call such a map a quasi-character,
%    reserving the word ``character'' for the conventional character of
%    absolute value $1$. Secondly, we shall find no model for the group
%    of quasi-characters, or even for the group of characters, though
%    such a model would be of the utmost importance.}
%  \cite{tate:67a}*{\S 2.3}
%\end{quotation}
%\noindent
In this paper we find a dictionary between quasicharacters of algebraic tori over local fields and quasicharacter sheaves. 
%
What is a quasicharacter sheaf?
Let $G$ be a smooth, commutative group scheme locally of finite
type over a finite field $\Fq$; $G$ need not be reductive, connected, or of
finite type over $\Fq$. Fix a prime $\ell$ invertible in $\Fq$
and write $m : \GxG \to G$ for the multiplication morphism,
$\bG$ for the base change of $G$ to $\bFq$ and
$\Frob{G} : \bG \to \bG$ for the Frobenius automorphism.
A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank~one and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms;
$\mu$ and $\phi$ are required to satisfy natural conditions explained below.

@@

We fix a prime $\ell$, invertible in $\Fq$.
We will work with constructible $\ell$-adic sheaves \cite{deligne:80a}*{\S 1.1} on $\QC$-schemes,
employing the standard formalism;
see, in particular, \cite{SGA5}*{Expos\'e VI,\S 2}, where pullback and tensor products are developed.
We also make extensive use of the external tensor product of $\ell$-adic sheaves, defined as follows: if $\mathcal{F}$ and $\mathcal{G}$ are constructible $\ell$-adic sheaves on schemes $X$ and $Y$, respectively, then $\mathcal{F}\boxtimes \mathcal{G} \ceq p_X^* \mathcal{F} \otimes p_Y^*\mathcal{G}$, where $p_X : X\times Y\to X$ and $p_Y : X\times Y \to Y$ are projections.

For any group $A$, we will denote by $A^*$ the dual group $\Hom_\text{grp}(A, \EEx)$.

@@

\begin{definition}\label{def:QC}
A \emph{quasicharacter sheaf on $G$} is a triple
$\qcs{L}\ceq (\gqcs{L},\mu,\phi)$ where:
\begin{enumerate}
\labitem{(CS.0)}{CS.0} $\gqcs{L}$ is a rank~one $\ell$-adic local system on $\bG$, by which we mean
a constructible $\ell$-adic sheaf on $\bG$, locally constant on each connected component of $\bG$, whose stalks are one-dimensional $\EE$-vector spaces; 
\labitem{(CS.1)}{CS.1} $\mu: \bm^* \gqcs{L} \to \LxL$ is an isomorphism of sheaves on $\bGxG$ such that the following diagram commutes, 
  where $m_3 \ceq m\circ (m\tight{1}{\times}{2}\id) = m\circ (\id\tight{2}{\times}{1} m)$;
%
  \[
  \begin{tikzcd}[row sep=30]
  \bm_3^*\gqcs{L} \arrow{rr}{(\bm \tight{1}{\times}{2} \id)^*\mu} \arrow[swap]{d}{(\id \tight{2}{\times}{1} \bm)^*\mu}
    &&  \bm^*\gqcs{L} \boxtimes \gqcs{L} \dar{\mu \tight{0}{\boxtimes}{1} \id} \\
    \gqcs{L} \boxtimes \bm^* \gqcs{L} \arrow{rr}{\id \boxtimes \mu}
    &&  \gqcs{L} \tight{0}{\boxtimes}{0} \LxL
  \end{tikzcd}
  \]
%
\labitem{(CS.2)}{CS.2} $\phi : \Frob{G}^* \gqcs{L} \to \gqcs{L}$ is an
  isomorphism of constructible $\ell$-adic sheaves on $\bG$ compatible with $\mu$ in the sense that the following diagram commutes.
%
  \[
  \begin{tikzcd}[row sep=20]
  \Frob{\GxxG}^* \bm^* \gqcs{L} \arrow{rr}{\Frob{\GxxG}^*\mu}
    && \Frob{\GxxG}^*(\LxL)\\
    \arrow[equal]{u} \bm^*  \Frob{G}^* \gqcs{L} \arrow[swap]{d}{\bm^* \phi}
    && \Frob{G}^*\gqcs{L}\boxtimes \Frob{G}^*\gqcs{L} \dar{\phi\boxtimes \phi} \arrow[equal]{u} \\
    \bm^*\gqcs{L} \arrow{rr}{\mu}
    && \LxL
  \end{tikzcd}
  \]
\end{enumerate}
\end{definition}

@@

\begin{proof}
  Let $\qcs{L}$ be a quasicharacter sheaf on $G$. We start by showing that
  $\bf^*\gqcs{L}$ is a local system of rank~one. Let $c_j :\bH^j \hookrightarrow \bH$ be any
  connected component and let $d_i : \bG^i \hookrightarrow \bG$ be the image of $j$ under the group
  homomorphism $\pi_0(\bf) : \pi_0(\bH) \to \pi_0(\bG)$.  Write $\bf^j : \bH^j \to \bG^i$
  for the restriction of $f$ to $\bH^j$.  Since $\gqcs{L}$ is locally constant on the connected
  components of $\bG$, there is a finite \'etale covering
  $a_i : X_i \to \bG^i$ such that $a_i^* (\gqcs{L}\vert_{\bG^i})$
  is constant.  Define $Y_j$ in the following diagram through pullback:
  \[
  \begin{tikzcd}[row sep=20, column sep=30]
   Y_j \dar[dashed]{g_j} \rar[dashed]{b_j} & \bH^j \dar{\bf^j} \rar[hook]{c_j} & \bH \dar{\bf} \\
   X_i \rar{a_i} & \bG^i \rar[hook]{d_i} & \bG
  \end{tikzcd}
  \]
  Then $b_j$ is a finite \'etale covering
  of $\bH^j$ and
  \[
  b_j^* \left( (\bf^* \gqcs{L})\vert_{\bH^j}\right)
  = (\bf\circ c_j \circ b_j)^*\gqcs{L}
  = (d_i\circ a_i\circ g_j)^*\gqcs{L} = g_j^* a_i^* (\gqcs{L}\vert_{\bG^i}).
  \]
%
  Since $a_i^* (\gqcs{L}\vert_{\bG^i})$ is a constant sheaf, its pullback
  $b_j^* \left( (\bf^*\gqcs{L})\vert_{\bH^j}\right)$ along $g_j$ is also constant.
  Thus $\bf^*\gqcs{L}$ is locally constant on
  $\bH^j$. To complete the proof that $\bf^*\gqcs{L}$ satisfies Condition~\ref{CS.0}
  simply observe that $\bf^*\gqcs{L}$ has rank~one since 
  $(\bf^*\gqcs{L})_{\bg} = \gqcs{L}_{f(\bg)}$, for every geometric point ${\bg}$ on $\bG$.

@@

\begin{definition}
Let $\QCb(G)$ be the category of pairs $(\qcs{L}_0,\mu_0)$ 
where $\qcs{L}_0$ an $\ell$-adic local system on $G$ of rank~one, 
equipped with an isomorphism $\mu_0 : m^* \qcs{L}_0 \to \qcs{L}_0 \boxtimes \qcs{L}_0$ 
satisfying the analogue of Condition~\ref{CS.1} on $G$; 
morphisms in $\QCb(G)$ are defined as in (the second part of) Condition~\ref{CS.2}. 
This is the category of \emph{bounded quasicharacter sheaves} on $G$. 
\end{definition}

@@

With Theorem~\ref{thm:snake} in hand, 
we can see why neither bounded quasicharacter sheaves 
nor finite quasicharacter sheaves are up to the task of 
geometrizing characters of $G(\Fq)$ in the generality we need.  
%
Consider the case when $G$ is the discrete \'etale group scheme $\ZZ$; 
this is indeed a $\QC$-scheme.
Let $\chi : G(\Fq) \to \EEx$ be the character determined by $\chi(1) = \ell$ 
and let $\qcs{L}$ be a quasicharacter sheaf in the isomorphism class 
corresponding to the character $\chi : G(\Fq) \to \EEx$ under Theorem~\ref{thm:snake}. 
Then $\trFrob{\qcs{L}}(1) = \ell$. Suppose $\qcs{L}$ was a bounded quasicharacter sheaf. 
Then the restriction of $\qcs{L}$ to the connected component $1\in G$ 
would be a $\ell$-adic local system on $1$ of rank~one, which would provide
an $\ell$-adic character $\rho_1 : \pi_1(1,{\bar 1}) \to \EEx$ with
$\rho_1(\Frob{\Fq}) = \ell$, necessarily continuous. 
\Clifton{I made some small changes to wording, here.}
But $\pi_1(1,{\bar 1}) = \Gal(\bFq/\Fq) \iso {\hat \ZZ}$ is compact, 
while $\rho_1(\pi_1(1,{\bar 1})) \subseteq \EEx$ is not compact (it contains $\ell^\ZZ$),
contradicting the continuity of $\rho_1$.
So the Weil sheaf $\gqcs{L}$ does not descend to a local system on $G$.
It follows that the quasicharacter sheaf $\qcs{L}$ is not a bounded quasicharacter sheaf.
%
We can play the same game with the character $\chi : G(\Fq) \to \EEx$ 
defined by $\chi(1) = 1+\ell$. In this case, the corresponding quasicharacter sheaf $\qcs{L}$ is bounded,
defined at $1$ by a pro-\'etale cover of $1$, but not finite, since $1+\ell$ is not a root of unity. 
%
These simple examples illustrate why the categories $\QCb(G)$ and $\QCf(G)$ are inadequate to geometrize characters of $G(\Fq)$.
